---
title: "Project 1"
author: "Name:  \n Partner: "
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}

#### Load necessary packages ####
# * These packages are not necessary to complete the assignment and or only used 
#   to provide an example. 
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)
```

## Background

The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases.
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*.
Your and your partner's role is to play a data scientist from one of these two entities.

## Data

> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths.
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user.
Date fields are stored in MM/DD/YYYY format.

## Project Objectives

### Objective 1

```{r ob1}
#stores the urls of the two data sets used in this objective 
time_series_covid19_confirmed_global <- ("time_series_covid19_confirmed_global.csv")
time_series_covid19_deaths_global <- ("time_series_covid19_deaths_global.csv")
#reads the two urls into variables and these csv's are stored as data frames
confirmedData <- read.csv(file=time_series_covid19_confirmed_global, header = TRUE)
deathData <- read.csv(file=time_series_covid19_deaths_global, header = TRUE)
#largest index stores the index of the country with the most deaths and confirmations on the first recorded day 
largestIndex <- 1
#scores the largest number of deaths and confirmations on the first day 
largestNumber <- 0
#Stores the current sum total of deaths and confirmations as we loop through the data sets. 
currentSum <- 0
#constant that stores the index of the first recorded day in the data set. 
INDEXOFFIRSTDAY <- 5

#for loop goes through every row of the data sets (since both data sets have the same dimension we can use nrow on either or)
for(i in 1:nrow(confirmedData)){
  #calculates the combined deaths and confirmations on the first day of the current row
  currentSum = (confirmedData[i,INDEXOFFIRSTDAY]) + (deathData[i,INDEXOFFIRSTDAY])
  #checks if the currentSum is larger than the previously found largest number 
  if((currentSum > largestNumber)){
    #sets the index of the largest number
    largestIndex = i
    #Sets the largest number as the currentSum for the next loop 
    largestNumber = currentSum
    
  }
  #resets the currentSum so that the loop can begin anew
  currentSum = 0

}
print(confirmedData[largestIndex,2])

```

### Objective 2

```{r ob2}

```

### Objective 3

```{r ob3}

```

### Objective 4

#### Objective 4.1

```{r ob4.1}

```

#### Objective 4.2

```{r ob4.2}

```

### GitHub Log

```{bash gitlog}
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```
